---
title: "rMSIproc analysis"
author: "Jordi Rofes"
date: "`r Sys.Date()`"
---

### Introduction
rMSI and rMSIproc are R packages that allow the visualization of mass spectometry imaging data (rMSI) and the following data processing from all the spectra for each pixel to a peak matrix (rMSIproc). This tutorial wants to provide a flexible and customizable workflow for scientists analyzing different rMSIproc peak matrixes obtained from different images. This tutorial will explain  how to: explore the peak matrix to visualize the different m/z intensities on the images, what normalization steps should be utilized when making comparisons, how to visualize the medium spectrum of the raw data, the application of multidimensional reduction methods (PCA), the utilization of clustering methods like Kmeans and SOM to find clusters of similar spectrums that can be related to differences between different tissue zones and the visualization and comparison of those clusters.

### Required Packages
For this tutorial you will need to install [rMSI](https://github.com/prafols/rMSI), [rMSIproc](https://github.com/prafols/rMSIproc), [ggplot2](https://github.com/tidyverse/ggplot2), [plotly](https://github.com/plotly/dashR) packages if you don't have them.

### rMSI

rMSI is a package that allows you to visualize mass spectometry imaging (MSI) images in an interactive. rMSi also allows to visualize and get the spectra of selected pixels and download them as CSV.  
```{r}
# This function opens the rMSI user interface that allows you to visualize a chosen image
rMSI::OpenMSI()
```
### rMSIproc

rMSiproc is a package that allows users to process MSI images by using the implemented user interface. The processing is followed by different steps: 
- Smoothing of the spectra
- Alignement of the spectra
- Mass calibration with known mass peaks: if checked this will require you to input a CSV with the known m/z, then an interface will allow you to asign these masses to the peak on the spectra and finally rMSI will move the m/z axis so that the chosen peak/s have the known m/z
- Peak peaking: peak peaking finds the peaks on the spectra and centroids them
- Mass binning: mass binning agroups different peaks on the same m/z so that all images have the same number of m/z.
After finishing rMSIproc will output a rMSIproc object with the peak matrix and some image information. 
```{r}
# This function opens the rMSIproc user interface.
# rMSIproc allows you to input multiple rMSI images to process them (peak peaking, binning, aligning, baseline correction, smoothing and calibration) and outputs a compressed file with the peak matrix
# Important: for this tutorial be sure all your images are processed at the same time and have the merge peak matrix option checked.
rMSIproc::ProcessWizard()

```

### First Steps

Before we begin working with the peak matrix we will need to tell R where all the files are located on your system. You will have to input the adress of your peak matrix, the name of your peak matrix and the location of the file initialPlotter.R
```{r message=FALSE, warning=FALSE}

# Insert here where your data is located (The peak matrix and the .tar images)
wDir <- "C:/Users/jordi/Desktop/Metabolomica/Projecte MALDI/CollCap/"
# Insert the name of the peak matrix .zip
peakData <- "mergeddata-peaks.zip"


library(rMSIflow) # This sources all the functions we will need
library(ggplot2)
library(plotly)
library(rMSIproc)


```

### Extracting the Peak Matrix

Now we can use the rMSIproc function LoadPeakMatrix to load the peak matrix into a variable.
This will create a rMSIproc object called peakM which will contain the peak matrix and useful information of the images (pixel position, names, SNR and area) 
```{r message=FALSE, warning=FALSE}

# Here we extract the Peak Matrix from the .zip and create and object with all the data assigned to peakM
peakM <- rMSIproc::LoadPeakMatrix(file.path(wDir, peakData))


# If your peak matrix is composed of multiple images this will separate them in multiple peak matrix
matrixList <- sepMatrix(peakM)

peakM$names 
```

### Normalizations
Before we visualize the data it's important to understand the methods of normalization you can use to normalize the data. 
In MSI the normalization step required is case-dependant but here we will explain all the normalization methods included on rMSIproc
processing. These normalization are included in the peak matrix object more exactly at peakM$normalizations and there is a
normalization value for each pixel on the image.

Data normalization is a key step to remove part of the systematical error so the data represents more the biological variation
  There are several ways to normalize the data:
     - TIC (Total Ion Count): TIC divides every pixel by the mean of all mass intesities of the raw data.
     - AcqTic (Acquired TIC): Acquired TIC calculates the TIC of each pixel and then calculates a mean with the surrounding TICs. 
     - RMS (Root Mean Squares): RMS divides each pixel by the root mean square of all mass intensities. 
     - Max: this method divides every pixel by the intensity it's most intense peak 

### Exploring the peak matrix

Now that we've seen the normalization methods we can proceed to the visualization of the data. This chunk below allows users to plot
m/z intensities of each pixel onto a heatmap with the for of the tissue. You can also choose the images to be plotted by using vectors
c(1:6), the normalization parameter can be changed by giving a vector with the normalization values for each pixel
(peakM$normalizations includes the normalizations rMSI proc calculated).     
```{r echo=FALSE, warning=FALSE}

# Select m/z to plot and image number out of peakM$names to plot
peakM$names
mz2plot <- 100            # You can find all the mz in your data using peakM$mass
img2plot <- 1:2           # Writing a vector c() allows you to plot more images at the same time. Input a 0 to plot all the images 
normalization <- peakM$normalizations$TIC  # Input here a vector with the normalization values. Input NA for no normalization.
savePlot <- F                 # Choose if you want to save the plots into a pdf file (T/F)
fileN <- "plotsRAW.pdf"   # Choose the name of the pdf file. If a file has the same name it will be overwritten


initialPlotter(peakM, matrixList, mz2plot,img2plot, normalization, savePlot, fileN)
```

### Medium Spectrum from raw data

For more information we can visualize the raw data by using rMSI functions to calculate de medium spectra of each image in profile
format. The medium spectrum is the spectrum that contains the mean of each m/z. The next first chunk will calculate the medium spectra
from all images and the second one allows users to compare them.
```{r}
# We can use rMSI functions to plot the average spectra of an image to explore the raw data
# Make sure that the .tar of the images is on the same directory you are working on
#medSpec <- medSpecRaw(wDir)

```

```{r message=FALSE, warning=FALSE}
# Here we plot the average spectrums we want to compare
avSpecimg2plot <- c(1:6)      # Select the images you want to compare. Writing c(1,2,3) allows you to plot more than 
                            # one image at the same time

library(ggplot2)
library(plotly)
medSpecRawplot(medSpec , avSpecimg2plot)

```

###  Class vector

Before we begin analyzing the data we will need to input a class vector. The class vector is a vector c() that contains the different experimental groups where the images belong. This vector can be useful when visualizing the images to be able to compare the different experimental groups. 
```{r}
# To allow classification of the images in different groups you will need to insert the names of each group in the vector groupsImg in the order they are on the peak matrix

peakM$names # This allows you to see the order of the images on the peak matrix

# Insert the group for each image (control, control, problem, ...) in the order they are on the peak matrix
# If you input NA. Each group will have the name in peakM$names
# If you want multiple images to have the same group you can simply write the group name twice
groupsImg <- c("Tumor1", "Tumor2", "Healthy1", "Healthy2", "Transition1", "Transition2")

paste0(rep("Control",10),1:10) #If you have many images per experimental group you may want to use an expression like this


if(length(groupsImg) != length(peakM$names)){
  stop("Be careful. The number of groups is not equal to the number of images. Every image should have a group assigned")
}

```

### Order changing

This next chunk is only useful if you want to change the order of the images on the peak matrix. It can be skipped if your images are already ordered.
```{r}
# If you are not happy with the order of the images showed on peakM$names you can change it's order by entering here a  
# vector with the desired order of images. This will update the peakM object with the new order

newOrder <- c(1,2,5,6,3,4) # Insert here the new desired order and run the code chunk. It will require some time to finish

matrixList <- matrixList[newOrder]
peakM <- rMSIproc::MergePeakMatrices(matrixList)
```

### Medium spectrum from peak matrix from selected pixels

This next chunk allows users to compare the medium spectra of chosen pixels. This medium spectrum is made calculating the mean of each mass from given pixels. This is useful for comparing different known zones that should have differend medium spectrum.
```{r}
# This function makes the medium spectrum from selectec pixels from the peak matrix

pixels2compare <- c(1:50)   # Input the first set of pixels that are going to be compared
pixels2compare2 <- c(51:100)# Input the second set of pixels that are going to be compared 
                            
name1 <- "Cortex"           # Insert a tag for the first set of pixels
name2 <- "Medula"           # Insert a tag for the second set of pixels
normalization <- peakM$normalizations$TIC    # Input here a vector with the normalization values. Input NA for no normalization.
savePlot <- F               # Choose if you want to save the plots into a pdf file (T/F)
fileName1 <- ".pdf"         # Choose the name of the pdf file. If a file has the same name it will be overwritten

medSpecPlot(peakM, pixels2compare, pixels2compare2, normalization, name1, name2, savePlot, fileName1)


```

### PCA

Principal component analysis (PCA) is a multidimensional reduction method. PCA computes a new set of variables from the originals rotating the axis so that the variance between the variables is maximized. This new set corresponds to the first principal component (PC1). Then the perpendicular axis from PC1 is calculated and the variables are again computed, this will correspond the second principal component (PC2). This can be done multiple times, so that we can calculate any amount of PCs we want.
```{r echo=FALSE, message=FALSE, warning=FALSE}
centeringJ <- T      # The centering ensures the data is found within the 0/0 origin
scalingJ  <- T       # The scaling equalizes the data variance to decrease the bias given the natural difference in units of the variables 
normalization <- peakM$normalizations$TIC  # Input here a vector with the normalization values. Input NA for no normalization.

pcaD <-pcaJr(peakM, centeringJ, scalingJ, normalization)

```

### Explained Variance

The explained variance is the percentatge of variance of each PC from the total variance of all PCs. Each PC has a percentatge of variance it explains from the original data and it decreses with each PC. This chunk creates a graphic that shows the cumulative percentatge of variance from all PC. This allows users to choose the number of principal components they should choose to explore.
```{r}
percentatge <- 0.7  # This marks the principal component with the cumulative percentatge of variance chosen from 0 to 1
  
pcChooseNum(pcaD, percentatge)


```


### PC vs PC plots

This chunk calculates the classic PC vs PC two dimensional graphic to visualize how well the PC separates the original data.
```{r}
pc2plot1 <- 1  # Select which PC are going to be plotted in a two dimensional graphic
pc2plot2 <- 2  # 
savePlot <- F  # Select if you want to save the plot into a pdf file (T/F)
fileNpca2 <- "PC1vsPC2.pdf" # Name of the .pdf file

plotPca <- pca2dimPlot(peakM, pcaD, groupsImg, pc2plot1,pc2plot2, savePlot, fileNpca2)
plotly::ggplotly(plotPca)

```

### PC plots on image

This chunk plots a chosen principal component onto the images.  
```{r message=FALSE, warning=FALSE}
pc2plot <- 1             # Select which PC to plot into the images
img2plotPca <- c(1,3)    # Select which images to plot. Using c() allows you to plot more images at the same time. 
savePlot <- F            # Save the plot into a pdf file (T/F)
fileNpca3 <- "PCimg.pdf" # Name of the .pdf file

pcaPlotImg(peakM, matrixList, pcaD, img2plotPca, groupsImg, pc2plot, savePlot, fileNpca3) 
  
```


### K-means

K-means is a clustering method. K-means generates random centers and associates each point to each center based on the euclidean distance to the center. Then the K-mean re-calculates the centers with the new associated points and repeats the process till the points associated to a center remain the same. Then the process is repeated with new centers and the clustering with the lowest cluster variance is chosen. 

First of all we will need to choose the number of clusters we will separate our data. For this task we will use the factoextra function fviz_nbclus which allows to use different methods for choosing the adequate number of cluster the data wants.
 - The elbow method (WSS): this method utilizes the within cluster sum of squares (WSS) which represents the intra-cluster variation.
The WSS is then calculated for each number of clusters (1, 2, 3,...10). The user should choose the number of clusters so that when adding another cluster it doesn't reduce as much the WSS. (The lower elbow part of the plot)
 - The average silhouette method: this method tries to mesure the quality of the clustering calculating how well each point lies within it's cluster. The user should choose the number of clusters that maximizes the average silhouette.
 - The gap method: this method compares the WSS with their expected values under null reference distribution of the data. The user should choose the number of clusters that maximizes the gap stadistic so that it means that the distribution is far from a randomly distributed

For now the only recommended method is the elbow method. The next chunk allow users to make an elbow plot and decide the number of clusters they will choose when making the k-means.
```{r}
data2clus <- peakM$intensity     # The data used to make the plots. It must be a matrix NxM where N are the observations and M the variables
norm <- peakM$normalizations$TIC # The normalization vector applied to the data
testNumClus <- 100               # The number of clusters to test. (Usually no more than 10 clusters are needed)
merged <- F                      # If the images have to be clustered together or separately

elbowPlot <- elbowMethod(peakM, data2clus, norm, testNumClus, merged)
ggplotly(elbowPlot)
# Silhouette <- factoextra::fviz_nbclust(peakM$intensity, kmeans, method = "silhouette")
# Silhouette
# Gap <- factoextra::fviz_nbclust(peakM$intensity, kmeans,  method = "gap_stat")
# Gap


```

```{r warning=FALSE}
data2clus <- peakM$intensity        # This variable allows you to choose the data you want to cluster. (It has to be a matrix of NxM 
                                    # dimensions where N must have the same length as the number of pixels of the peak matrix)
norm4 <- peakM$normalizations$TIC   # Normalization applied to the data
numClusters <- 4                    # Number of cluster to make the k-means
merged <- F                         # This variable allows you to choose between clustering all images together (T) or clustering them
                                    # separately (F)
kmeansData <- kmeansCluster(peakM, data2clus, norm4, numClusters, merged)

```


### K-means segmentation plots

This next chunk plots the cluster vector onto the images. This allows users to see which pixels correspond to each cluster.
```{r}
img2plot <- c(1,4)    # Images you want to plot. 
savePlot <- F         # Save the plot into a pdf file (T/F)   
fileName3 <- ".pdf"   # Name of the .pdf file

ggplotly(clusterDataPlotting(peakM, matrixList, img2plot, groupsImg, kmeansData, savePlot, fileName3))


```


### K-mean cluster comparison

This chunk allows users to calculate the medium spectra of chosen clusters and images. This allows users to visualize the spectra differences of each cluster and image. 
Remember that if the K-means is done with all the images at the same time (merge == T) there may be images that don't contain any pixel from one of the clusters. These clusters won't have a medium spectrum associated with the image.
```{r warning=FALSE}
img2plot <- c(1,4)               # Choose the images you want to plot
clus2comp <- c(3,4)              # Choose the clusters you want to plot
norm <- peakM$normalizations$TIC # Input here a vector with the normalization values. Input NA for no normalization.

clusterKmeansComparison(peakM, norm, img2plot, groupsImg, clus2comp, kmeansData)

```

### Comparing Clusters

After comparing medium spectra from clusters we will want to compare the differences between the medium spectra of some of the clusters.
In the next chunk you can choose the image and cluster from that image you want as the reference spectrum and those you want as the spectrum to compare. This will output a list with a data frame with four columns: 
- The mz compared
- The log2(fold change)
- The p-value from a t-test corrected by fdr (false discovery rate)
- The error estimation (ppm error from closest gold peak found)
```{r}
# Remember if an image doesn't have the chosen cluster it won't work
refImg <- 1     # Choose the image where the reference cluster is located
refClus <- 3    # Choose the cluster you want to compare as reference
compImg <- 4    # Choose the image where the cluster you want to compare is located
compClus <- 3   # Choose the cluster from the image you want to use as comparison 
norm <- peakM$normalizations$TIC  # Input here a vector with the normalization values. Input NA for no normalization.

sumDataComp <- compareClusMedSpec(c(refImg, refClus), c(compImg, compClus), peakM, kmeansData, norm)
```

### Comparing other spectra

Sometimes you might not want to compare the clusters of one image. Other options might be interesting for analysis like comparing two full images. For this reason you can use this chunk to obtain the same results as the chunk before but with the chosen pixels of the spectra. Note that the functions input 

```{r}
refpixels <- c(1:400)   # The pixels that contain the reference spectrum
comppixels <- c(401:600)# The pixels that contain the spectrum to compare
norm <- peakM$normalizations$TIC # Input here a vector with the normalization values. Input NA for no normalization.

data2comp <- medSpecComp(peakM$intensity[refpixels,], peakM$intensity[comppixels,], peakM, norm)
```

### Volcano Plot

This next chunk creates a volcano plot with x = -log2(Fold Change) and y = -log10(p-values) and some lines marking the p-value = -log10(0.05) and the Fold Change = +-log2(2).
```{r}
data2plot <- sumDataComp # The data frame obtained from the comparison of spectra

plotVolcano <-volcanoPlotJ(data2plot)
ggplotly(plotVolcano)

```

### Next steps

After comparing the different spectra the next step would be to search for compounds on databases. rMSIproc last version contains a function for annotating peak matrixes called peakAnnotation. This function searches for monoisotopics m/z and adducts and annotates them.

Note that this function is only on the latest rMSIproc version and is still in development. This means that it can sometimes cause R to crash. Be sure to save your workspace before using it. If it doesn't work you should reestart R and retry. If that doesn't work then the m/z have to be annotated manually.
```{r}
peakMant <- peakAnnotation(peakM)
```
These m/z can be used to search on databases and then assign them to a list of putative compounds.





