---
title: "rMSIproc analysis"
author: "Jordi Rofes"
date: "`r Sys.Date()`"
output: html_document
---

### Introduction



### Preparations

```{r}

# Insert where your data is located (The peak matrix and the .tar images)
wDir <- "C:/Users/jordi/Desktop/Metabolomica/Projecte MALDI/CollCap/" 

# Insert the name of the peak matrix .zip
peakData <- "mergeddata-peaks.zip"

# Insert the group for each image (control, control, problem, ...) in order
groupsImg <- c("Control", "Control", "Transition", "Transition", "Tumor", "Tumor") 

```

### Extracting the Peak Matrix

```{r results = "hide", echo = "false", include=matrixSeparation}

# Here we extract the Peak Matrix from the .zip and create and object with all the data assigned to peakM
peakM <- rMSIproc::LoadPeakMatrix(file.path(wDir, peakData))

# If your peak matrix is composed of multiple images this will separate them in multiple peak matrix
if(length(peakM$numPixels) != 1){
  matrixList <- matrixSeparation(peakM)
}

```

### Exploring the peak matrix

```{r warnings= "hide"}

#Select m/z to plot and image number out of peakM$names to plot
peakM$names
mz2plot <- 600       #Tip# You can find all the mz in your data at peakM$mass
img2plot <- c(4,6,5)   #Tip# Writing a vector c() allows you to plot more images at the same time

plots <- initialPlotter(peakM, matrixList, mz2plot,img2plot, groupsImg)

pl <- lapply(1:length(img2plot), function(i){
  plotly::ggplotly(p = plots[[i]])
})
plotly::subplot(pl)



```

### Normalizations

Data normalization is a key step to remove part of the systematical error so the data represents more the biological variation
  There are several ways to normalize the data:
     TIC (Total Ion Count): 
     aqTIC (aquired TIC)
     RMS ()

Altres normalitzacions que podem probar ---> Mediana // TIC excluint pics molt o poc intensos // Nivell de soroll?¿ -> Que la variancia sigui la mateixa  ->fer diferencies consecutives entre punts dels espectres i després buscar la mediana de -> cada diferencia menys la mediana

```{r}
mz <- 400
rMSIproc::plotPeakImage(matrixList[[1]], mz, labels = groupsImg[1])
rMSIproc::plotPeakImage(matrixList[[1]], mz, labels = groupsImg[1],normalization = "TIC")
rMSIproc::plotPeakImage(matrixList[[1]], mz, labels = groupsImg[1],normalization = "RMS")
rMSIproc::plotPeakImage(matrixList[[1]], mz, labels = groupsImg[1],normalization = "AcqTic")

#pl <- plotly::ggplotly(rMSIproc::plotPeakImageG(matrixList[[1]], mz, plot_labels = groupsImg[1]))
#pl2 <- plotly::ggplotly(rMSIproc::plotPeakImageG(matrixList[[1]], mz, plot_labels = groupsImg[1], normalization = "TIC"))
#plotly::subplot(pl, pl2)

```

### Medium Spectrum

```{r}
# We can use rMSI functions to plot the average spectra of an image to explore the raw data
# First you will need to put the .tar of the images on the same directory you are working on
bad <- list.files(wDir, pattern = "^ramdisk")
imageName <- list.files(wDir, pattern = "proc.tar")
imageName <- imageName[!(imageName %in% bad)]

# Then we run the code so it extracts the image information and calculates de average spectrum
avSpec <- lapply(imageName, function(name){
  image <- rMSI::LoadMsiData(file.path(wDir, name))
  mass <- image$mass
  aS <- rMSIproc::AverageSpectrum(image) 
  spect <- cbind(mass, aS)
  return(spect)
})

avSepcData <- data.frame(avSpec)
row.names(avSepcData) <- imageName
colnames(avSepcData) <- image$mass

```

```{r}
# Here we plot the average spectrums we want to compare


plotly::ggplotly(ggplot2::ggplot(avSepcData) + geom_line(aes(x = mass, y = aS), colour = "red")+
                   geom_line(aes(x = avSepcData$mass, y = avSepcData$aS.1),colour = "blue")+
                   geom_line(aes(x = avSepcData$mass, y = avSepcData$aS.2),colour = "yellow")+
                   geom_line(aes(x = avSepcData$mass, y = avSepcData$aS.3),colour = "green")+
                   geom_line(aes(x = avSepcData$mass, y = avSepcData$aS.4),colour = "orange")+
                   geom_line(aes(x = avSepcData$mass, y = avSepcData$aS.5),colour = "black")+ xlab("M/Z") + ylab("Intensity"))
#rMSI::plotSpectra(mass = image$mass, intensity = avSpec1)
```



### PCA

```{r }

dataNormTic <- peakM$intensity/peakM$normalizations$TIC ### Dividim les intensitats pel factor de normalització TIC 
pca <- prcomp(dataNormTic, center = T, scale. = T) ### Fem la PCA de la matriu de pics

for (i in which(pca$sdev > 2)){
  
  rMSIproc::plotValuesImage(peakM, pca$x[,i])

}

```

### K-means

```{r warning=FALSE}

dataNormTic <- peakM$intensity/peakM$normalizations$TIC ### Dividim les intensitats pel factor de normalització TIC 

clData <- vector(mode = "list",length = length(peakM$numPixels)) 

for (i in 1:length(peakM$numPixels)){ 
 if (i == 1){
    clData[i] <- list(kmeans(dataNormTic[1:peakM$numPixels[i], ], 2))
 } else {
    clData[i] <- list(kmeans(dataNormTic[(sum(peakM$numPixels[1:(i-1)])+1):sum(peakM$numPixels[1:i]), ], 2))
 }
}

clusImages <- lapply(clData, function(x){
  return(x[1]) 
})

allClusImages <- unlist(clusImages)


rMSIproc::plotValuesImage(peakM, allClusImages)

```

### Plots PCA + K-means

```{r}


```


package.skeleton(name = initialPlotter,list = initialPlotter.R, path = C:/Users/jordi/Desktop/Metabolomica/Projecte MALDI/initialPlotter)

